
% Software Sustainability Challenge submission should be no more than
% 2-pages long, and use the ACM sigconf template (for more details,
% see the main conference Call For Papers). They will be peer
% reviewed, and accepted submissions will be presented at the
% conference as either part of a panel or lightning talk and/or within
% a poster session (as determined by the Programme Chair). Challenge
% submissions will not be included in the main DLfM proceedings, but
% will be published on the conference website; they will additionally
% be used (with appropriate credit given) to inform a report into
% Digital Musicology for the UK Software Sustainability Institute.
%
% Challenge contributions must be submitted via the DLfM 2023
% EasyChair page by Friday 29 September.

\documentclass[sigconf]{acmart}
%\documentclass[sigconf, nonacm=true, authordraft=true, anonymous=true]{acmart}
\setcopyright{none}

\begin{document}
\title{Software Sustainability Challenge: ECOLM and Lute Tablature}

\acmConference[DLfM'23]{10th International Conference on Digital Libraries for Musicology}{November 10, 2023}{Milan, Italy}

\author{Chris Cannam}
\orcid{https://orcid.org/0009-0001-9814-6512}
\affiliation{%
  \institution{Particular Programs Ltd}
  \city{London}
  \country{UK}}
\email{chris.cannam@particularprograms.co.uk}

\author{David Lewis}
\affiliation{%
  \institution{Goldsmiths, University of London}
  \city{London}
  \country{UK}}
\email{d.lewis@gold.ac.uk}

\author{Tim Crawford}
\affiliation{%
  \institution{Goldsmiths, University of London}
  \city{London}
  \country{UK}}
\email{t.crawford@gold.ac.uk}

\maketitle
\begin{sloppypar}

  % Authors should use headings from the following list to structure
  % their submissions, addressing implications for sustainability in all
  % relevant sections, and using as many as are applicable for their
  % scenario:
  %
  % ·       Nature and purpose of the sustainable tool or resource
  % - Yes
  %
  % ·       Audience and users, disciplines and subjects
  % - Yes
  %
  % ·       Position within the research lifecycle
  % - Possibly not
  %
  % ·       Means of access and accessibility
  % - Possibly not
  %
  % ·       Current and future requirements, or implementations
  % - Yes
  %
  % ·       Challenges for sustainability
  % - Yes, especially
  %
  % ·       Future directions
  % - Yes, tentatively

  \section{Nature and purpose of the sustainable tool or resource}

  {\bf ECOLM} (1999-2002) was a project led by Tim Crawford which
  developed and populated a database of lute tablature encodings with
  metadata, for scholarly use, queried using a web
  interface. Subsequent projects {\bf ECOLM II} (2002-2006) and {\bf
    ECOLM III} (2012) expanded the database and used it for some
  computational musicological investigations. The resulting database
  was hosted on a public-facing web server at Goldsmiths, University
  of London.\footnote{http://doc.gold.ac.uk/isms/ecolm/database/} It
  is still running today, although nobody is formally responsible for
  maintaining it.

  ECOLM is a relatively small scholarly resource with around 2,000
  carefully-curated encodings. A number of other public lute tablature
  resources exist, of varying size, quality, and consistency (see
  section \ref{audience}). These typically face challenges to
  sustainability similar to those of ECOLM (see section
  \ref{challenges}). It seems useful to consider those challenges for
  such resources in general, at least insofar as they address the same
  audience.
  
  \section{Audience and users, disciplines and subjects}\label{audience}

  We are considering here the sustainability of at least five lute
  tablature datasets: ECOLM itself; {\bf
    lutemusic.org}\footnote{https://lutemusic.org/} by Sarge Gerbode;
  {\bf mss.slweiss.de}\footnote{https://mss.slweiss.de/} curated by
  Peter Steur and the late Markus Lutz; a set of scans from {\bf Lute
    Society publications} curated by John Robinson; and a set of
  transcriptions of {\bf editions by Pierre Phal\`ese} curated by Jan
  Burgers. All of these are either Creative Commons licensed or have
  been offered by their maintainers as possible constituents of a
  future combined tablature resource.

  These resources serve a spectrum of audiences. At one end,
  lutemusic.org aims at performers and includes edited transcriptions
  with relatively little scholarly metadata or editorial comment. At
  the other, ECOLM is aimed at computational musicologists and
  prioritises diplomatic facsimiles and transcriptions that preserve
  original scribal idiosyncracies.

  In this review we are particularly interested in sustainability for
  musicology and other academic purposes. To this end, we asked three
  exemplary users of online early-music resources---a musicologist, a
  computational musicologist, and a lute teacher---for their views
  about them in order to understand scholarly expectations.

  Briefly, they agreed on the importance of trust and provenance,
  particularly in knowing about the quality of transcription and level
  of editorial intervention in a resource. There was some consensus
  about the value of simple search with subsequent refinement, of
  cleanly laid-out results including inline incipits, and of API and
  data provision. Resources mentioned as worth learning from included
  DIAMM\footnote{https://www.diamm.ac.uk/} (Digital Image Archive of
  Medieval Music), the Vihuela
  Database,\footnote{https://vihuelagriffiths.com/} the Josquin
  Research Project,\footnote{https://josquin.stanford.edu/} and
  RISM\footnote{https://rism.info/} (R\'epertoire International des
  Sources Musicales) which is a near-ubiquitous entry point for
  musicological queries.
  
  \section{Challenges for sustainability}\label{challenges}

  \section{Future directions}

  We have identified three alternative directions for sustainable
  development.
  
  \subsection{``Enhanced ECOLM''}

  This approach would retain the relational data schema of the
  existing ECOLM, which is detailed and fairly effective, although it
  has little in common with the other systems we have considered or
  with wider current practice. We then provide ETL-type (extract,
  transform, load) data loaders for other sources of interest.

  Advantages of this approach include the ability to preserve existing
  code and to use original ECOLM records as a reference. The existing
  schema provides appropriate structure and reflects some good
  domain-specific decisions. Relational data import is a well
  understood field, and we could focus on user interfaces and data
  conversion rather than any novelty of data representation.

  Disadvantages include that the schema has little in common with any
  of the ad-hoc solutions other maintainers have settled on, so all
  import and export would be custom. The schema is perhaps already
  overspecified for its current use, yet does not address any problems
  relating to stable identifiers, versioning, or providing queryable
  APIs or data sources.

  Although we could at least initially reuse the existing user
  interface, it is no longer considered a strength of the system and
  would need some work to update to modern expectations.

  \subsection{Graph-based}

  In this approach we would take the fundamental representation to be
  a graph of triples in the model of RDF, and convert all metadata to
  that for import and from it for query. External data such as
  transcriptions and multimedia resources are identified by
  graph-relatable identifiers such as URIs.

  Advantages include the use of a widely-understood and accepted model
  that meets common expectations about data compatibility and API
  provision. For schema we can draw ontologies from a number of
  existing systems. The structure is reasonably amenable to versioning
  and to use of ``idempotent'' import flows with automated testing,
  offering the option of ongoing import of changes in upstream
  sources. In principle existing tools may be used for review, query,
  inferencing, and format conversion.

  The approach has difficulties as well. It discards the existing user
  interface work and requires even the existing ECOLM data to be
  converted. Although graph representations have wide application,
  they are not generally used for manual data management and therefore
  have as little in common with the ad-hoc schemas of enthusiast lute
  resources as with that of ECOLM. Significant work would be required
  to maintain stable identifier mappings from external sources. With a
  more flexible structure than ECOLM's relational database, care and
  good automated testing would be needed to avoid ``silently missing
  data'' problems on query. Finally a separate solution would be
  needed to the problem of identifying and retrieving non-graph data
  such as media resources.

  Although in this approach we could no longer use the existing ECOLM
  user interface, that may be slightly mitigated by the ability to
  adapt other graph-driven UIs to the model.
  
  \subsection{``RISM-aligned''}

  Fundamental metadata representation is MARC, and we use Muscat to
  maintain it? 
  
\end{sloppypar}
\end{document}
